---

title: "LIFE748 Assessment 2 - Part 2: Machine Learning Analysis of Differential Expression"

---

# GAI Declaration

**Tool used:** ChatGPT (OpenAI)

**Purpose of use:** Assistance with R code for differential expression analysis, unsupervised clustering, supervised classification, debugging, and interpretation.

**Prompts used:**\
1. "Help me perform PCA, k-means clustering and hierarchical clustering on DEG expression data in R."\
2. "How do I run logistic regression, LDA and SVM on RNA-seq DEG data in R?"\
3. "How do I evaluate model performance using confusion matrices, sensitivity and specificity in R?"

# Introduction

This analysis investigates whether machine learning approaches can discriminate between two experimental conditions using *E. coli* gene expression data. The workflow includes differential expression analysis, feature selection, unsupervised clustering, and supervised classification.

# Step 0: Load Required Libraries

```{r}
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(caret)
library(MASS)
library(e1071)
```

# Step 1: Load the Count Matrix and Define Sample Metadata

```{r}
counts <- read.csv("dataset_LIFE748_2.csv", row.names = 1)

colData <- data.frame(
  row.names = colnames(counts),
  Condition = c(rep("CondA", 25), rep("CondB", 25))
)

dim(counts)
table(colData$Condition)
head(counts[, 1:6])
```

# Step 2: Differential Expression Analysis with DESeq2

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = colData,
  design = ~ Condition
)

dds <- DESeq(dds)

res <- results(dds)
res <- as.data.frame(res)

head(res)
summary(res)
```

# Step 3: Variance-Stabilising Transformation

```{r}
vsd <- vst(dds, blind = FALSE)
normalized_counts <- assay(vsd)

dim(normalized_counts)
```

# Step 4: Select Differentially Expressed Genes

Relaxed thresholds were used to retain enough informative features for machine learning.

```{r}
res_clean <- res[!is.na(res$padj), ]

deg_genes <- rownames(
  res_clean[res_clean$padj < 0.1 & abs(res_clean$log2FoldChange) > 0.5, ]
)

length(deg_genes)
head(deg_genes)
```

# Step 5: Prepare the Machine Learning Dataset

The DEG matrix is transposed so that samples are rows and genes are columns.

```{r}
deg_matrix <- normalized_counts[deg_genes, ]
ml_data <- t(deg_matrix)
ml_data <- as.data.frame(ml_data)
ml_data$Condition <- colData$Condition

dim(ml_data)
head(ml_data[, c(1:5, ncol(ml_data))])
```

# Step 6: Unsupervised Learning

## 6.1 Principal Component Analysis (PCA)

```{r}
pca_res <- prcomp(ml_data[, -ncol(ml_data)], scale. = TRUE)

pca_df <- data.frame(
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2],
  Condition = ml_data$Condition
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 3) +
  theme_minimal() +
  ggtitle("PCA of DEG Expression Profiles")
```

## 6.2 K-means Clustering

```{r}
set.seed(123)

km_res <- kmeans(ml_data[, -ncol(ml_data)], centers = 2, nstart = 25)

table(Cluster = km_res$cluster, Condition = ml_data$Condition)
```

## 6.3 Hierarchical Clustering Heatmap

```{r}
annotation_df <- data.frame(
  Condition = ml_data$Condition
)
rownames(annotation_df) <- rownames(ml_data)

pheatmap(
  t(ml_data[, -ncol(ml_data)]),
  annotation_col = annotation_df,
  show_colnames = FALSE,
  main = "Heatmap of Differentially Expressed Genes"
)
```

# Step 7: Supervised Learning

## 7.1 Train-Test Split

```{r}
set.seed(123)

train_index <- createDataPartition(ml_data$Condition, p = 0.7, list = FALSE)

train_data <- ml_data[train_index, ]
test_data  <- ml_data[-train_index, ]

train_data$Condition <- factor(train_data$Condition)
test_data$Condition  <- factor(test_data$Condition)

table(train_data$Condition)
table(test_data$Condition)
```

## 7.2 Logistic Regression

```{r}
lr_model <- glm(
  Condition ~ .,
  data = train_data,
  family = "binomial"
)

lr_prob <- predict(lr_model, test_data, type = "response")
lr_pred <- ifelse(lr_prob > 0.5, "CondB", "CondA")
lr_pred <- factor(lr_pred, levels = levels(test_data$Condition))

lr_cm <- confusionMatrix(lr_pred, test_data$Condition)
lr_cm
```

## 7.3 Linear Discriminant Analysis (LDA)

```{r}
lda_model <- lda(Condition ~ ., data = train_data)

lda_pred <- predict(lda_model, test_data)
lda_class <- lda_pred$class

lda_cm <- confusionMatrix(lda_class, test_data$Condition)
lda_cm
```

## 7.4 Support Vector Machine (SVM)

```{r}
svm_model <- svm(Condition ~ ., data = train_data, kernel = "linear")

svm_pred <- predict(svm_model, test_data)

svm_cm <- confusionMatrix(svm_pred, test_data$Condition)
svm_cm
```

# Step 8: Compare Classifier Performance

```{r}
model_results <- data.frame(
  Model = c("Logistic Regression", "LDA", "SVM"),
  Accuracy = c(
    as.numeric(lr_cm$overall["Accuracy"]),
    as.numeric(lda_cm$overall["Accuracy"]),
    as.numeric(svm_cm$overall["Accuracy"])
  ),
  Sensitivity = c(
    as.numeric(lr_cm$byClass["Sensitivity"]),
    as.numeric(lda_cm$byClass["Sensitivity"]),
    as.numeric(svm_cm$byClass["Sensitivity"])
  ),
  Specificity = c(
    as.numeric(lr_cm$byClass["Specificity"]),
    as.numeric(lda_cm$byClass["Specificity"]),
    as.numeric(svm_cm$byClass["Specificity"])
  )
)

model_results
```

# Step 9: Cross-Validation (Recommended Extension)

A single train-test split can be sensitive to random partitioning. The following code performs 5-fold cross-validation for LDA and SVM to provide a more robust performance estimate.

```{r}
set.seed(123)

ctrl <- trainControl(method = "cv", number = 5)

cv_lda <- train(
  Condition ~ .,
  data = ml_data,
  method = "lda",
  trControl = ctrl
)

cv_svm <- train(
  Condition ~ .,
  data = ml_data,
  method = "svmLinear",
  trControl = ctrl
)

cv_lda
cv_svm
```
